---
output: 
  html_document:
    keep_md: true
---

```{r, echo=FALSE}
# Initial Notes 

# Assignment
# 
#   The basic goal of this assignment is to explore the NOAA Storm Database and 
#   answer some basic questions about severe weather events. 

#   You must use the database to answer the questions below and
#   show the code for your entire analysis. 

#   Your analysis can consist of tables, figures, or other summaries. 

#   You may use any R package you want to support your analysis.
# 
# Questions
# 
#   Your data analysis must address the following questions:
# 
#   1.Across the United States, which types of events (as indicated in the EVTYPE variable) 
#   are most harmful with respect to population health?
# 
# 
#   2.Across the United States, which types of events have the greatest 
#   economic consequences?
# 
# 
# Consider writing your report as if it were to be read by a government or 
# municipal manager who might be responsible for preparing for severe weather 
# events and will need to prioritize resources for different types of events. 
# However, there is no need to make any specific recommendations in your report.

# Document Layout

#   -Language: Your document should be written in English.
#   -Title: Your document should have a title that briefly summarizes your data analysis
#   -Synopsis: Immediately after the title, there should be a synopsis which describes 
#     and summarizes your analysis in at most 10 complete sentences.
#   -There should be a section titled Data Processing which describes (in words and code) 
#     how the data were loaded into R and processed for analysis. 
#     In particular, your analysis must start from the raw CSV file containing the data. 
#     You cannot do any preprocessing outside the document. 
#     If preprocessing is time-consuming you may consider using the cache = TRUE option 
#     for certain code chunks.
#   -There should be a section titled Results in which your results are presented.
#   -You may have other sections in your analysis, but 
#     Data Processing and Results are required.
#   -The analysis document must have at least one figure containing a plot.
#   -Your analyis must have no more than three figures. 
#     Figures may have multiple plots in them (i.e. panel plots), but 
#     there cannot be more than three figures total.
#   -You must show all your code for the work in your analysis document. 
#     This may make the document a bit verbose, but that is okay. 
#     In general, you should ensure that echo = TRUE for every code chunk 
#     (this is the default setting in knitr).


```

# Title:  Assessment of NOAA Storm Data and its Impact on Population Health and Economic Consequences 

## Synopsis:  
 
In this analysis, I look at NOAA Storm Data from 1950 to 2011.  I explore the impact of storms and severe weather and assess their impact on both the health of individuals in the population at large in addition to their impact on the economies of the areas those storms occurred.  Primarily I looked at 2 measures {property damage, crop damage} to assess economic impact and 2 measures {fatalities, injuries} to assess health impact.  An interesting outcome is that the answer varies based on which measure you choose.  For example, the top Event is Flood for Property Damage, Drought for Crop Damage, Tornado for Fatalities, and Tornado for Injuries.  Even though 2 of the events line up as a top measure, the 2nd most impactful event for each measure do not line up for Fatalities and Injuries.  Also to note, I interpreted "figure" as being a "plot", else I would have commented out more of my Exploratory Data Analysis.  The plots are at the end of the document in the results section.  

## Data Processing:  

Storm Data is obtained, per the Assignment, from the course website <a href='https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2'>here</a>, however per the instructors, the data originated from the NOAA.  

Additional documenation related to this data is also provided, both related to its <a href='https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf'>dataset preparation</a> and a <a href='https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf'>FAQ</a>.  

Next, I read in the dataset, and do some exploratory data analysis on the data set. Note, as the Assignment required no more than 3 Plots, the Exploratory Plots are either commented out or removed (with some left here as representative of some of the plots / Exploratory Data Analysis done).    

After reading in the data, the PROPDMGEXP was used to translate the PROPDMG into Actual $ with the result stored in PROPDMGACT.  Similarly, the CROPDMGEXP was used to translate the CROPDMG into Actual $ with the result stored in CROPDMGACT.  These two computed variables, PROPDMGACT and CROPDMGACT, in addition to the two additional original variables, FATALITIES and INJURIES, are used here for analysis.  

Since the Assignment wants to know the the impact of Storms, both in terms of economic impact and personal health, I use the two computed variables, PROPDMGACT and CROPDMGACT, as a proxy to the Economic Damage Impact.  Likewise, I use the two additional original variables, FATALITIES and INJURIES, as a proxy to the Personal Health Impact.  

Next, the Assignment requested the impact to be determined by EVTYPE, and as such, I summarize the weather data into an aggregate dataset where summary statistics{sum, mean, sd, median, min, max} were computed for each variable {PROPDMGACT, CROPDMGACT, FATALITIES, INJURIES}.  

Please note, some of the preprocessing, e.g. the "head(weatherData.orig)" and the "unique(weatherData.orig$EVTYPE)", were commented out to to the sheer # of pages they added to the printout.  Since both of these were done on the raw dataset to which everyone had access, they were determined the best options to minimize the # of pages in the writeup (when originally completed, I had over 40 pp, with 10pp due just to the unique call).  

The Data Preprocessing code is as follows:  

```{r, echo=TRUE}

# read in data from file 
weatherData.orig <- read.table(
    file = "..\\..\\Data\\repdata_data_StormData.csv.bz2", 
    header = TRUE, sep = ",", na.strings = "NA", nrows = 2500000 # 250000
  )
    #colClasses = c("numeric", "factor", "factor")

# basc summary statistics 
#head(weatherData.orig)
summary(weatherData.orig)
dim(weatherData.orig)

# focus on impact to population and economy - identify the variables and provide basic statistcs 
quantile(weatherData.orig$FATALITIES)
quantile(weatherData.orig$INJURIES)
quantile(weatherData.orig$PROPDMG)
quantile(weatherData.orig$CROPDMG)

# unique(weatherData.orig$EVTYPE)
unique(weatherData.orig$PROPDMGEXP)
unique(weatherData.orig$CROPDMGEXP)

library(dplyr)

# filter for "B" and get subset 
weatherData.PROP.B <- weatherData.orig %>% 
  select(everything()) %>% 
  filter(PROPDMGEXP %in% c("B")) 
  
# filter for "M" and get subset 
weatherData.PROP.M <- weatherData.orig %>% 
  select(everything()) %>% 
  filter(PROPDMGEXP %in% c("M")) 
  
# filter for "M" and get subset 
weatherData.PROP.K <- weatherData.orig %>% 
  select(everything()) %>% 
  filter(PROPDMGEXP %in% c("K")) 
  
# filter for "M" and get subset 
weatherData.PROP.N <- weatherData.orig %>% 
  select(everything()) %>% 
  filter(!(PROPDMGEXP %in% c("B", "M", "K"))) 
  
# add Property Damage in Actual $ to the respective datasets 
weatherData.PROP.B <- cbind(weatherData.PROP.B, PROPDMGACT = weatherData.PROP.B$PROPDMG * (1000 * 1000 * 1000))
weatherData.PROP.M <- cbind(weatherData.PROP.M, PROPDMGACT = weatherData.PROP.M$PROPDMG * (1000 * 1000))
weatherData.PROP.K <- cbind(weatherData.PROP.K, PROPDMGACT = weatherData.PROP.K$PROPDMG * (1000))
weatherData.PROP.N <- cbind(weatherData.PROP.N, PROPDMGACT = weatherData.PROP.N$PROPDMG * (1))

# resssemble dataset 
weatherData.INTERIM <- rbind(weatherData.PROP.B, weatherData.PROP.M, weatherData.PROP.K, weatherData.PROP.N) 

# filter for "B" and get subset 
weatherData.CROP.B <- weatherData.INTERIM %>% 
  select(everything()) %>% 
  filter(CROPDMGEXP %in% c("B")) 
  
# filter for "M" and get subset 
weatherData.CROP.M <- weatherData.INTERIM %>% 
  select(everything()) %>% 
  filter(CROPDMGEXP %in% c("M")) 
  
# filter for "M" and get subset 
weatherData.CROP.K <- weatherData.INTERIM %>% 
  select(everything()) %>% 
  filter(CROPDMGEXP %in% c("K")) 
  
# filter for "M" and get subset 
weatherData.CROP.N <- weatherData.INTERIM %>% 
  select(everything()) %>% 
  filter(!(CROPDMGEXP %in% c("B", "M", "K"))) 

# add CROP Damage in Actual $ to the respective datasets 
weatherData.CROP.B <- cbind(weatherData.CROP.B, CROPDMGACT = weatherData.CROP.B$CROPDMG * (1000 * 1000 * 1000))
weatherData.CROP.M <- cbind(weatherData.CROP.M, CROPDMGACT = weatherData.CROP.M$CROPDMG * (1000 * 1000))
weatherData.CROP.K <- cbind(weatherData.CROP.K, CROPDMGACT = weatherData.CROP.K$CROPDMG * (1000))
weatherData.CROP.N <- cbind(weatherData.CROP.N, CROPDMGACT = weatherData.CROP.N$CROPDMG * (1))

# resssemble dataset 
weatherData.ALL <- rbind(weatherData.CROP.B, weatherData.CROP.M, weatherData.CROP.K, weatherData.CROP.N) 

# group by EVTYPE 
weatherData.EVTYPE <- weatherData.ALL %>% 
  group_by(EVTYPE) %>% 
  summarize(
            sumPROPDMGACT = sum(PROPDMGACT, na.rm = TRUE), 
            meanPROPDMGACT = mean(PROPDMGACT, na.rm = TRUE), 
            sdPROPDMGACT = sd(PROPDMGACT, na.rm = TRUE), 
            medianPROPDMGACT = median(PROPDMGACT, na.rm = TRUE), 
            minPROPDMGACT = min(PROPDMGACT, na.rm = TRUE), 
            maxPROPDMGACT = max(PROPDMGACT, na.rm = TRUE),
            
            sumCROPDMGACT = sum(CROPDMGACT, na.rm = TRUE), 
            meanCROPDMGACT = mean(CROPDMGACT, na.rm = TRUE), 
            sdCROPDMGACT = sd(CROPDMGACT, na.rm = TRUE), 
            medianCROPDMGACT = median(CROPDMGACT, na.rm = TRUE), 
            minCROPDMGACT = min(CROPDMGACT, na.rm = TRUE), 
            maxCROPDMGACT = max(CROPDMGACT, na.rm = TRUE), 

            sumFATALITIES = sum(FATALITIES, na.rm = TRUE), 
            meanFATALITIES = mean(FATALITIES, na.rm = TRUE), 
            sdFATALITIES = sd(FATALITIES, na.rm = TRUE), 
            medianFATALITIES = median(FATALITIES, na.rm = TRUE), 
            minFATALITIES = min(FATALITIES, na.rm = TRUE), 
            maxFATALITIES = max(FATALITIES, na.rm = TRUE), 

            sumINJURIES = sum(INJURIES, na.rm = TRUE), 
            meanINJURIES = mean(INJURIES, na.rm = TRUE), 
            sdINJURIES = sd(INJURIES, na.rm = TRUE), 
            medianINJURIES = median(INJURIES, na.rm = TRUE), 
            minINJURIES = min(INJURIES, na.rm = TRUE), 
            maxINJURIES = max(INJURIES, na.rm = TRUE) 
            
            )

# explore results  
head(weatherData.EVTYPE)
summary(weatherData.EVTYPE)

```

## Results:  

Above, I did a tremendous amount of preprocessing as already documented.  Here, the results are presented.  Also to note, I interpreted "figure" as being a "plot" (in the question also, it implies figure is synonymous with plot(s), not data tables), else I would have commented out more of my Exploratory Data Analysis.  

First, I look at the aggregate weather data sorted by PROPDMGACT, CROPDMGACT, FATALITIES, and INJURIES, each of which grouped by EVTYPE, in order to see which EVTYPE (Storm Type) has the biggest impact in each of these 4 measures.  

```{r, echo=TRUE}
## 3 figures - can use multi-panel plots 

# sort by sumPROPDMGACT 
weatherData.EVTYPE.BySumPROPDMGACT <- weatherData.EVTYPE %>%
    select(EVTYPE, sumPROPDMGACT) %>% 
    arrange(desc(sumPROPDMGACT)) 
head(weatherData.EVTYPE.BySumPROPDMGACT)

# sort by sumCROPDMGACT 
weatherData.EVTYPE.BySumCROPDMGACT <- weatherData.EVTYPE %>%
    select(EVTYPE, sumCROPDMGACT) %>% 
    arrange(desc(sumCROPDMGACT)) 
head(weatherData.EVTYPE.BySumCROPDMGACT)

# sort by sumFATALITIES 
weatherData.EVTYPE.BySumFATALITIES <- weatherData.EVTYPE %>%
    select(EVTYPE, sumFATALITIES) %>% 
    arrange(desc(sumFATALITIES)) 
head(weatherData.EVTYPE.BySumFATALITIES)

# sort by sumINJURIES 
weatherData.EVTYPE.BySumINJURIES <- weatherData.EVTYPE %>%
    select(EVTYPE, sumINJURIES) %>% 
    arrange(desc(sumINJURIES)) 
head(weatherData.EVTYPE.BySumINJURIES)

```

From the above, in terms of Property Damage, the biggest impact storm types (in descending order) are:  Flood, Hurricane / Typhoon, and Tornado.  Similarly, in terms of Crop Damage, the biggest impact storm types (in descending order) are:  Drought, Flood, and River Flood.  Further, in terms of Fatalities, the biggest impact storm types (in descending order) are:  Tornado, Excessive Heat, and Flash Flood.  Lastly, in terms of Injuries, the biggest impact storm types (in descending order) are:  Tornado, Thunderstorm Wind, and Flood.  As mentioned earlier, the answer to what is the most harmful to Economic Health or Public Health depends on which measure you choose.   

Next, looking at the data graphically the aggregate weather data sorted by PROPDMGACT, CROPDMGACT, FATALITIES, and INJURIES, each of which grouped by EVTYPE, in order to see which EVTYPE (Storm Type) has the biggest impact in each of these 4 measures.  

```{r, echo=TRUE}
## 3 figures - can use multi-panel plots 

# Top5 Unique values by Impact Measure 
Top5.BySumPROPDMGACT = head(weatherData.EVTYPE.BySumPROPDMGACT, 5)
Top5.BySumCROPDMGACT = head(weatherData.EVTYPE.BySumCROPDMGACT, 5)
Top5.BySumFATALITIES = head(weatherData.EVTYPE.BySumFATALITIES, 5)
Top5.BySumINJURIES = head(weatherData.EVTYPE.BySumINJURIES, 5)

# Plot Only Top 10 
par(mfcol = c(2,2), mar = c(4,4,2,1))

barplot(height = Top5.BySumPROPDMGACT$sumPROPDMGACT, 
  names.arg = Top5.BySumPROPDMGACT$EVTYPE, 
  main="Economic - Property Damage By Event Type", 
  xlab="Event Type",
  ylab="Economic - Property Damage ($)"
  ) 
  
barplot(height = Top5.BySumCROPDMGACT$sumCROPDMGACT, 
  names.arg = Top5.BySumCROPDMGACT$EVTYPE, 
  main="Economic - Crop Damage By Event Type", 
  xlab="Event Type",
  ylab="Economic - Crop Damage ($)"
  ) 

barplot(height = Top5.BySumFATALITIES$sumFATALITIES, 
  names.arg = Top5.BySumFATALITIES$EVTYPE, 
  main="Personal - Fatalities By Event Type", 
  xlab="Event Type",
  ylab="Personal - Fatalities (#)"
  ) 

barplot(height = Top5.BySumINJURIES$sumINJURIES, 
  names.arg = Top5.BySumINJURIES$EVTYPE, 
  main="Personal - Injuries By Event Type", 
  xlab="Event Type",
  ylab="Personal - Injuries (#)"
  ) 

```

Above, we look graphically at the aggregate weather data sorted by PROPDMGACT, CROPDMGACT, FATALITIES, and INJURIES, each of which grouped by EVTYPE, in order to see which EVTYPE (Storm Type) has the biggest impact in each of these 4 measures.  Again, in terms of Property Damage, the biggest impact storm types (in descending order) are:  Flood, Hurricane / Typhoon, and Tornado.  Similarly, in terms of Crop Damage, the biggest impact storm types (in descending order) are:  Drought, Flood, and River Flood.  Further, in terms of Fatalities, the biggest impact storm types (in descending order) are:  Tornado, Excessive Heat, and Flash Flood.  Lastly, in terms of Injuries, the biggest impact storm types (in descending order) are:  Tornado, Thunderstorm Wind, and Flood.  As mentioned earlier, the answer to what is the most harmful to Economic Health or Public Health depends on which measure you choose.   

In summary, I looked at the NOAA Storm Data from 1950 to 2011.  I explored the impact of storms and severe weather and assessed their impact on both the health of individuals in the population at large in addition to their impact on the economies of the areas those storms occurred.  Primarily I looked at 2 measures {property damage, crop damage} to assess economic impact and 2 measures {fatalities, injuries} to assess health impact.  An interesting outcome is that the answer varies based on which measure you choose, with Flood, Drought, or Tornado being some of the top events, depending on the measure of damage / impact.  
